Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-08-27 review/Response to reviews.Rmd

##  {#R1-5}

> I suggest that a little more about the limitations of this modeling approach in the discussion would help set the context for interpreting the results and make the paper more robust.

<!--# The following isn't in the main text yet.  Just ideas. -->

(Something I'm considering adding to the discussion):

"The main limitation for detecting lagged effects with any method is data, and the specifics of data limitations have not yet been systematically explored. @tellerLinkingDemographyDrivers2016 used simulations to show that 20--25 years of data were required to detect lagged effects using their method, but this cutoff depends on the effect size and the number of climate extremes in the dataset. Additionally, there is some evidence that appropriate spatial replication could drastically reduce the number of years of data required (Aldo Compagnoni's ESA talk)."

@@

We thank the reviewer for pointing out this mistake.
We have edited this line to read "Although plants were significantly larger in continuous forest compared to fragments, the effect of size in year t on size in year t+1 is nearly identical in the two habitats." See also our response to \@ref(R1-figs).

##  {#R2-6}

> Lines 287-289.
> While this statement is true, there are also years where flowering is lower in CF (e.g., 2002 similar values as following the drought in 2003 in fragments).
> I feel that the reporting of the results is a bit biased towards concluding that plants do worse in fragments compared to CF, especially since none of the result reported so far are significant.

<!--# need to double-check and revise -->

The difference in flowering probability between cointinuous forests and fragments is in fact significant.
We have added Table 1 that shows the model intercepts ± 84% confidence intervals, which is comparable to testing a main effect of habitat if we were able to include both habitats in the same model.

@@

...see also our response to \@ref(R1-5).
<!--# R1-5 is asking for more discussion on limitations of DLNMs -->

##  {#R2-14}

> 424-427: I think this is exaggerating since the differences between habitat types is not significant.

<!--# the magnitude *was* greater, interaction not tested -->

...see also our response to \@ref(R2-10)

### 2021-08-27 doc/paper.Rmd
update results section with new results.  Closes #72

where $y_{ij}$ is the $i$th plant observation in the $j$th plot $s_1(z_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details).
To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
A random effect of plant ID was not included in survival models since each plant only dies once.
Individual-level random effects were not included due to computational limitations in fitting random intercepts for all of the nearly 5,000 plants in continuous forest plots.
Preliminary modeling with data subsets showed that excluding individual-level random effects reduced model R^2^, but did not affect the shape or statistical significance of the smooths.

@@

## Survival, growth, and flowering in continuous forest vs. fragments

@@

*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.93$ in both continuous forest and forest fragments).
This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (i.e. negative growth in Figure \@ref(fig:eda)a).
When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; Table \@ref(tab:intercept-table)).
However, survival in both habitats was significantly affected by size in the previous year (Table \@ref(tab:results-table)) with larger plants having much greater survival probabilities (Figure \@ref(fig:size-plot)b).
The survival probability of large plants approached 1 in both habitat types, but the smallest plants had higher survival in 1-ha fragments.

@@

*Size:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 39.8 cm tall (± 26.2 SD).
Plants in 1-ha fragments had on average `r format((1 - (2.45/2.85))*100, digits = 3)`% fewer shoots (2.5 ± 1.4 SD) and were `r format((1 - (35.57/39.75))*100, digits = 3)`% shorter (35.6 cm ± 23.8 SD).
Because our proxy for plant size was the product of these two metrics, plants in continuous forest were on average 32% larger than those in forest fragments (144 ± 176 SD vs. 109 ± 139 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
Plants were significantly larger in continuous forest (Table \@ref(tab:intercept-table)), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time (Figure \@ref(fig:eda)a).

Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
Although plants were significantly larger in continuous forest compared to fragments, the effect of size in year t on size in year t+1 is nearly identical in the two habitats (Figure \@ref(fig:size-plot)a).

@@

*Flowering:* The overall proportion of plants that flowered was very low.
Flowering probability was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively); this difference was statistically significant (Table \@ref(tab:intercept-table)).
The observed disparity was largely due to the fact that flowering is also significantly size-dependent (Table \@ref(tab:results-table)), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
Despite the flowering probability of the largest plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
The most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. 17% in 1-ha fragments (Figure \@ref(fig:eda)c).

@@

*Survival:* For 1-ha fragments, there was a significant effect on survival of SPEI in the preceding 13 months.
The highest survival was near SPEI of 0, with mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
Wet conditions in the preceding 23 months (i.e., SPEI \> 1) also had a significant negative effect on survival in fragments (Figure \@ref(fig:surv-spei)b).
In contrast, the effect of recent SPEI in continuous forest was weaker, with only the 6 months preceding a census having a significant effect on survival (Figure \@ref(fig:surv-spei)a).
These short-term effects of SPEI on survival were also unidirectional---the probability of survival declined, albeit only slightly, with increasingly negative values of SPEI (i.e., as droughts became more severe; Figure \@ref(fig:surv-spei)a).
In contrast, the most pronounced negative effects of SPEI on the survival of plants in continuous forest were at lag times of 15--20 months and 32--36 months.
Drought 15--20 months prior to a census (i.e. two dry seasons prior to a census) was significantly associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was significantly associated with higher survival.
Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
It should be noted, however, that only the first transition year of census data (1998--1999) met these conditions.
We compared the effects of SPEI history in continuous forest and fragments by subtracting the fitted values in Figure \@ref(fig:surv-spei)b from Figure \@ref(fig:surv-spei)a to produce Figure \@ref(fig:surv-spei)c.
This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.025.

*Growth:* The effects of drought history on trends in plant size were generally similar for continuous forest and fragments.
However, the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)) <!--# this is no longer true.  need to update -->.
For example, under average conditions (i.e., SPEI = 0), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
However, the growth of plants in fragments is reduced when the current wet season is unusually wet (i.e., SPEI\>2), and as a result the plants in continuous forests will be larger by up to log(size) = 0.57.
In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, with a more pronounced response in 1-ha fragments.

*Flowering:* Overall, the the probability of flowering was significantly higher in continuous forest than in 1-ha fragments (Figure \@ref(fig:flwr-spei), Table \@ref(tab:intercept-table)).
SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure \@ref(fig:flwr-spei)b).
This led to some important inter-habitat differences in plant responses to prior droughts.
In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1), drought two dry seasons prior (lags 15--20) and in the wet season 34--36 months prior all increased the probability of flowering.
However, drought at the end of the rainy season one year prior (lags 7--13) significantly reduced flowering probability (Figure \@ref(fig:flwr-spei)a).
In fragments, recent drought had no significant effect on flowering probability and only drought at two dry seasons prior (lags 16--20) and in the wet season 33--36 months prior increased flowering probability (Figure \@ref(fig:flwr-spei)b).
The effects of drought on flowering probability were strong in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were on average more likely to be larger and flower again (Table \@ref(tab:results-table)).
Finally, with the exception of the model for survival in 1-ha fragments, the random effect of plot was significant, indicating that vital rates varied among plots (Table \@ref(tab:results-table)).

@@

Finally, any analytical approach for assessing lagged effects on demography requires long-term data and/or data collected from many sites with independent weather (but similar climate and site conditions) [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021] <!--# add citation to Aldo's ESA talk -->.
Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.
More generally, however, researchers need to consider how delayed responses to climate could influence the interpretation of data in studies where the organisms lifespan exceeds the study's duration.

>>>>

where $y_{ij}$ is the $i$th plant observation in the $j$th plot $s_1(z_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details).
To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
A random effect of plant ID was not included in survival models since each plant only dies once.
Individual-level random effects were not included due to computational limitations in fitting random intercepts for all of the nearly 5,000 plants in continuous forest plots.
Fitting models with a data subset with and without individual-level random effects showed that excluding individual-level random effects reduced model R^2^, but did not affect the shape or statistical significance of the smooths.

@@

## Survival, size, and flowering in continuous forest vs. fragments

@@

*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year (93% survival in both continuous forest and forest fragments).
This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (i.e. negative growth in Figure \@ref(fig:eda)a).
There was no significant difference in survival between continuous forest and 1 ha fragments (Table \@ref(tab:intercept-table)).
Survival in both habitats was significantly affected by size in the previous year (Table \@ref(tab:results-table)) with larger plants having much greater survival probabilities (Figure \@ref(fig:size-plot)b).
The survival probability of large plants approached 1 in both habitat types, but the smallest plants had higher survival in 1 ha fragments (although 95% confidence intervals overlap at all sizes).

@@

*Size:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 39.8 cm tall (± 26.2 SD).
Plants in 1-ha fragments had on average `r format((1 - (2.45/2.85))*100, digits = 3)`% fewer shoots (2.5 ± 1.4 SD) and were `r format((1 - (35.57/39.75))*100, digits = 3)`% shorter (35.6 cm ± 23.8 SD).
Because our proxy for plant size was the product of these two metrics, plants in continuous forest were on average 32% larger than those in forest fragments (144 ± 176 SD vs. 109 ± 139 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
Plants were significantly larger in continuous forest (Table \@ref(tab:intercept-table)), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time.

Mean plant size dropped dramatically in 2003 in both habitat types (negative fold-change in Figure \@ref(fig:eda)a), corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
As with survival, size in year t was a significant predictor of size in year t+1 (Table \@ref(tab:results-table)).
Although plants were significantly larger in continuous forest compared to fragments (Table \@ref(tab:intercept-table)), the effect of size in year t on size in year t+1 is nearly identical in the two habitats (Figure \@ref(fig:size-plot)a).

@@

*Flowering:* The overall proportion of plants that flowered was very low.
The proportion of plants flowering was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]*100`% vs. `r summary_stats$flwr_mean[1]*100`%, respectively).
The intercepts for the continuous forest and 1 ha fragment models were significantly different, indicating a main effect of habitat on flowering probability (Table \@ref(tab:intercept-table)).
The observed disparity in proportion of flowering plants was largely due to the fact that flowering is also significantly size-dependent (Table \@ref(tab:results-table)), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
Despite the flowering probability of the largest plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
The most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. 17% in 1-ha fragments (Figure \@ref(fig:eda)c).

@@

*Survival:* For 1-ha fragments, there was a significant effect on survival of SPEI in the preceding 0--16 months.
The highest survival was near SPEI of 0, with mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
In contrast, only recent conditions (0--4 month delay) has a significant effect on survival in continuous forest (Figure \@ref(fig:surv-spei)a).
The short-term effects of SPEI on survival in continuous forest were also unidirectional---the probability of survival was highest in wet conditions and declined, albeit only slightly, with increasingly negative values of SPEI (i.e., as droughts became more severe; Figure \@ref(fig:surv-spei)a).
Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
It should be noted, however, that only the first transition year of census data (1998--1999) met these conditions.
We compared the effects of SPEI history in continuous forest and fragments by subtracting the fitted values in Figure \@ref(fig:surv-spei)b from Figure \@ref(fig:surv-spei)a to produce Figure \@ref(fig:surv-spei)c.
This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.042.

*Size:* The effects of drought history on trends in plant size were generally similar for continuous forest and fragments.
Under all conditions, plant size is greater in continuous forest (Figure \@ref(fig:growth-spei)c).
Decreasing SPEI at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats.
In continuous forest, but not 1 ha fragments, SPEI at lags of 22--25 months had a significant effect on plant size in the opposite direction, with wet conditions at that lag time resulting in larger predicted sizes.
At longer lags of 26--36 months positive effects of drought on plant size are predicted for both continuous forest and 1 ha fragments by our models, although the area that reaches statistical significance is larger for continuous forest.

*Flowering:* Overall, the probability of flowering was significantly higher in continuous forest than in 1-ha fragments (Figure \@ref(fig:flwr-spei), Table \@ref(tab:intercept-table)).
Recent and past SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure \@ref(fig:flwr-spei)b).
This led to some important inter-habitat differences in plant responses to prior droughts.
In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1) and droughts two dry seasons prior (lags 15--20) increased the probability of flowering.
The shape of the crossbasis smooth for 1 ha fragments suggested that moderate drought (-1.5 \<SPEI\< -1) in the previous 0--18 months and at lags of 28--36 months significantly increased flowering probability slightly (the highest probability estimated is 0.004; Figure \@ref(fig:flwr-spei)b).
The effects of drought on flowering probability appeared stronger in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were significantly more likely to be larger and flower again.
Finally, with the exception of the model for survival in 1-ha fragments, the random effect of plot was significant, indicating that vital rates varied among plots (Table \@ref(tab:results-table)).

@@

Finally, any analytical approach for assessing lagged effects on demography requires long-term data and/or data collected from many sites with independent weather (but similar climate and site conditions) [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021] <!--# add citation to Aldo's ESA talk -->. <!--# possibly good place to mention limitations of interactions -->Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.
More generally, however, researchers need to consider how delayed responses to climate could influence the interpretation of data in studies where the organisms lifespan exceeds the study's duration.
